Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl

### 2023-09-27 Code/Draft_ruddy_duck_vital_rates.Rmd
revison A. Caizergues

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]
- Jean-François Maillard^[Office Français de la Biodiversité, France - jean-francois.maillard@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]
date: "`r Sys.setlocale('LC_TIME', 'English'); format(Sys.time(), '%d %B, %Y')`"
abstract: |
  Monitoring the number of individuals is by far the most popular strategy for investigating the environmental factors ruling the population dynamics and for measuring the efficacy of management actions targeting population recovery, control or eradication. Unfortunately, it is insufficient for the basic understanding of the demographic mechanisms and more specifically to assess the extend to which population growth rate is affected by changes in adult survival rather than to variations in reproductive parameters. Usual methods to inform survival and recruitment (capture-mark-recapture, game-hunting bag) rely on catching animals in a way or another, which suffers from three main drawbacks. First, the selectivity of the catching methods can lead to a biased representation of the underlying population structure. In practice, catching and releasing animals is also an invasive approach implying additional stress to the individuals. Furthermore releasing may simply be forbidden for invasive alien species. In many waterfowl species, a sexual dichromatism is observed in adults whereas immatures of both sexes display a cover-up plumage similar to adult females. From populations of Ruddy duck, a species displaying this pattern, this study introduces a non-invasive method based on count data to estimate the respective contribution of recruitment and adult survival on population growth rate. Relying on matricial models shows that performing count survey during the appropriate time window to differentiate both plumage types is sufficient to provide long-term time series of the two main demographic components of the population growth rate.

  **Keywords**: reproductive success - fecundity - productivity - age ratio - vital rate - duck - Oxyura jamaicensis

output: 
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
    bookdown::word_document2:
      toc: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

The Ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in Great Britain in the 40’s [@GutierrezExposito2020], a feral population began to establish with the first observed reproductive attempts in the wild in the 60s (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5,000 individuals in the early 2000s. By the end of the 80s, a feral population began to establish also on the continent in France, supposedly owing to the arrival of individuals born in Great Britain. Contrary to what was observed in Great Britain, the French Ruddy duck population did not spread much, with the vast majority of observations and breeding attempts clustering in Western France. In France, during winter, almost no Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which is thus the exclusive wintering site and greatly facilitates the monitoring of this population. The feral European populations of Ruddy duck are considered as a major threat to the endangered native White-headed duck, *Oxyura leucocephala*, population of the South-Western Mediterranean, because of hybridization risks and thus an elevated potential for genetic pollution and eventually genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, eradication measures were taken both in Great Britain and in France from the late 90s [@GutierrezExposito2020], (Figure \@ref(fig:count)), and a European Ruddy duck eradication plan has been initiated by the European Commission in 1999 [@Hughes1999]. The release of non-native species in the wild, including Ruddy ducks, is forbidden in countries of the European Union, and the status of White-headed duck populations is highly unfavourable. The use of capture-mark-recapture to monitor populations of these species is thus impossible. As a result, the effectiveness of the eradication of Ruddy ducks and management actions intended to favour the recovery of White-headed ducks were essentially assessed through censuses. Censuses, however, prevented assessment of the relative effects or changes in survival and recruitment rate on population growth rate, which is a key towards a proper understanding of the efficiency of management actions.

@@

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism. Thereby immature males acquire the typical colourful plumage of their kind over the course of the interbreeding season, which is typically in late winter. Immature males are thus confounded to immature and adult females before their moult (Figure \@ref(fig:flock)). Delayed dichromatism explains why the apparent proportion of females in the winter counts is always greater than the true proportion of females that can be assessed in removals (Figure \@ref(fig:pfem)). Thus, it is possible to assess the proportion of immatures within a population by monitoring the seasonal evolution of apparent sex ratio and assuming an even sex ratio at birth [@Bellrose1961; @Blums1996], comparable mortality rates between females and males among juveniles [@Ramula2018; @Wood2021], and unchanging adult sex ratio over the monitoring period [@Devineau2010]. Then, both adult survival rate and recruitment rate can be estimated by combining the age ratio thus obtained with reliable counts over two successive winters.

@@

In waterfowl, survival rate is commonly higher in adult males than in adult females [@Baldassarre2014]. Adult sex ratio is then generally biased towards males (Figure \@ref(fig:pfem), @Szekely2014). As a consequence, deducing the proportion of immatures within a population from the observed proportion of adult males is not straightforward because the proportion of adults is not just twice as much as the observed proportion of adult males. There were no counts differentiating male-like from female-like individuals during the breeding period to properly estimate adult sex ratio. But sex identification in adults from removals was available for both Ruddy duck populations. However, the small population size in France prevented from getting precise adult sex ratios. A comparison over months in Great Britain showed no difference in proportion of males in adults, so data from adults collected over the whole year were used to estimate the proportion of males in adults. A comparison of proportion of males in adults among years with more than 500 samples did not show significant interannual differences. Even if the adult sex ratio may vary over a long time range, it is relatively stable over a few years [@Wood2021]. As a consequence, removal data from all adults were pooled to estimate the proportion of males in adults, see Equation \@ref(eq:pm).

@@

|Name|Class|Description|
|:---|:----|:----------------------------|
|$AM$|Data|Total number of adult males in hunting bags|
|$AF$|Data|Total number of adult females in hunting bags|
|$p(m|a)$|Parameter|Proportion of males in adults (or probability of being a male knowing it is an adult)|
|$CML_{i, t}$|Data|Cumulated number of male-like individuals counted in population $i$ in year $t$, which are assumed to all be adult males|
|$C_{i, t}$|Data|Cumulated number of individuals counted in population $i$ in year $t$|
|$p(a \cap m)_{i, t}$|Parameter|Proportion of adult males in population $i$ in year $t$|
|$p(a)_{i, t}$|Parameter|Proportion of adults in population $i$ in year $t$|
|$p(i)_{i, t}$|Parameter|Proportion of immatures in population $i$ in year $t$|
|$N_{i, t}$|Data|Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population)|
|$s_{i, t}$|Parameter|Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$|
|$r_{i, t}$|Parameter|Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per individual in year $t - 1$|
|$S_{i, t}$|Parameter|Number of adults in population $i$ in year $t$|
|$R_{i, t}$|Parameter|Number of recruits in population $i$ in year $t$|

@@

The relevance of the method was assessed by testing the likelihood of the estimates of both vital rates. For adult survival rate, we checked if the values belonged to the expected interval $[0;1]$, and we compared the values to literature data on ducks that are accessible because adult survival is commonly assessed by capture-mark-recapture [@Lebreton2001]. Maximum adult survival is defined as a uniform distribution on $[0.7;1]$. The lower limit corresponds to the upper range of survival rates in literature for waterfowl species of similar weight, the upper limit was set to 1 because survival rates of long-life waterfowl species are very high [@Nichols1997; @Krementz1997; @Buxton2004; @Nichols1992]. Recruitment rate is more tricky to validate because it is not upper bounded as it is defined on $[0;\infty[$ and is neither explicitly informed in the literature. Only some components of the recruitment rate are generally described, such as nesting rate, nesting success, clutch size, hatching success, early stage survival, but rarely survival between the juvenile stage and the first breeding season, which is necessary to inform recruitment rate [e.g. @Baldassarre2014]. We then developed an approach to estimate the maximum expected recruitment rate without exploitation and without negative density-dependence processes, and we checked if all values were equal or below this maximum recruitment rate. If the recruitment rate outputs of the model are significantly higher than maximum recruitment rate, it means that the method overestimates such rate, which reflects that a part of adult males are not detected during count surveys. A null or negative recruitment rate would mean that immature males moulted before the survey or/and the assumption on the stability of the adult sex ratio is overrated.

@@

The method successfully provided estimates and associated uncertainty of immature proportion in both wintering populations. Adult survival rate and recruitment rate, i.e. the two components of the population growth rate, were also estimated by the method successfully(Figure \@ref(fig:vital)).

@@

The male proportion in adults, which is a prerequisite to infer on the immature proportion, is estimated at 0.60 [0.59; 0.61]. The immature proportion varies from 0.16 [0.07; 0.24] to 0.54 [0.44; 0.62]. The lower values are similar for the two populations, but the range of the immature proportions in the GB population is far more limited (upper value : 0.30 [0.14; 0.45]). Conversely from the FR population, there is no significant difference from a year to another because all estimates overlap each other. We can notice that the immature proportion for the GB population is quite stable over the time series available, but a slight decrease is observed for the FR population.

The adult survival rate varies from 0.33 [0.29; 0.37] to 1.01 [0.89; 1.14]. All estimates are not significantly outside the range of a survival rate defined without immigration [0; 1]. There is no trend in adult survival for both populations, but we can notice two main differences. The GB population has the lowest values of adult survival rate, and even if there are significant difference among years for both populations, the variability for the GB population is a bit lower.

The recruitment rate varies from 0.09 [0.04; 0.13] to 0.75 [0.63; 0.87]. All estimates are above 0, meaning there is no senseless estimation. The maximum recruitment rate being 0.68 [0.36; 0.78] (see Section \@ref(MAX)), all estimates are not significantly outside the range of a recruitment rate defined without immigration [0; 0.78]. Similarly to the immature proportion, there is no trend for the GB population, but a decrease is observed for the FR population even if it is more noisy than for the immature proportion. Similarly to the survival rate, the GB population has low values of recruitment rate, and even if there are significant difference among years for both populations, the variability for the GB population is far lower than for the FR population.

The lower variability of both vital rates for the GB population explains its steady population growth over the considered time series compared to the FR one (Figure \@ref(fig:count)). In the FR population, the range of the recruitment rate (min/max difference: 0.66) is higher than the range of the survival rate (min/max difference: 0.58). Conversely, the range of the survival rate (min/max difference: 0.46) in the GB population is far higher than the range of the recruitment rate (min/max difference: 0.08). This reflects that the variability of the population growth rate for the FR population is mostly driven by the recruitment rate whereas the main source of variability is the adult survival for the GB population.

@@

From the 5 years available in the GB population, immature proportions estimated from the count-based method is positively correlated to estimates from hunting-bag data (Figure \@ref(fig:cor)). The hypothesis that the proportion of female-like individuals in the wintering populations reflects the age structure of these populations is then supported by this result. This correlation is even stronger when considering vital rates (Figure \@ref(fig:cor)). This result is expected since the two methods estimating these vital rates have a component in common, the growth rate of the population (see Equation \@ref(eq:vr)).

It is noticeable that a 1:1 correlation is never obtained. Immature proportion is lower in counts than in hunting bags. This result is not surprising because immature individuals are expected to be more vulnerable to hunting than adults and might then be over represented in hunting bag data. As a consequence, the correlation coefficients for both vital rates differ also from 1. This difference between the two approaches does not question the previous validation of the ability of the count method to track the interannual variability of the vital rates.

@@

## Response of vital rates to the harvest strategy {#MAX}

Under no harvest pressure, the maximum growth rates of the two populations are very similar (Figure \@ref(fig:mgr)), respectively 1.45 [1.36; 1.55] and 1.52 [1.25; 1.88] for the GB and the FR population. It corresponds to an increase of 45% per year and 52% per year respectively. Based on literature, we assumed a maximum survival rate similar for both populations 0.85 [0.7; 1]. We deduced a maximum recruitment rate of 0.60 [0.42; 0.78] for the GB population and of 0.68 [0.36; 1.06] for the FR population.

Even if a significant part of culled individuals were not aged for some years, the proxies of the harvest pressure show significant shifts over time (Figure \@ref(fig:sup)). This led to consider a high harvest pressure period from 2006 with no clear difference between the two populations because the signal is very noisy. Even if the harvest pressure seems comparable, the harvest strategies differ significantly between the two populations, see Section \@ref(HARV). Under high harvest pressure, the growth rate drops to 0.56 [0.53; 0.59] and 0.97 [0.94; 1.00] for the GB and the FR population respectively, which corresponds to a decrease of 44% per year and 3% per year. This result shows that the GB population significantly decreases during the high harvest pressure whereas the FR population is stabilized.

The adult survival rate under high harvest pressure is 0.44 [0.42; 0.46] and 0.62 [0.59; 0.64] for the GB and FR populations respectively, when the recruitment rate decreases to 0.13 [0.11; 0.15] and 0.32 [0.29; 0.34] respectively. The drop of the growth rate of the GB population corresponds to a similar average drop in adult survival rate (-0.41) and in recruitment rate (-0.47). The stabilisation of the FR population is more due to a decrease of the recruitment rate (-0.36) than from a drop of the adult survival (-0.23). For both populations, the response to the harvest pressure is significant, meaning that the compensation of the population to harvest mortality is limited if not null. Vital rates of the GB population are more affected than the FR population. Unsurprisingly, the FR harvest strategy focusing on the postbreeding period affects mostly the recruits. However, the prebreeding harvest strategy applied on the GB population that targets adults performs to deplete strongly the recruitment to a very low level (0.13 recruits produced per breeder).

@@

The delayed sexual dichromatism in a waterfowl species is used here to develop and test the validity of a new non-invasive method aiming to break down the apparent growth of a population into its two structural components: the adult survival rate and the recruitment rate. Disentangling the contribution of these two vital rates is a key to understand the response of a population to management measures. The development of the method holds on a unique dataset of two comparable populations of a same species, the Ruddy duck. Both populations were tracked from their introduction in a similar environment to a period of intense harvest pressure. This common history trajectory enables to evaluate the maximum growth of this species, but also its response to harvest pressure. As the harvest strategy differs between the two populations, we detail the similarities and the discrepancies of the responses of the vital rates, and open the discussion to the generalization of this method and its implications for waterfowl management.

@@

The vital rate estimates of the two populations belongs to the same range, which demonstrate a certain consistency in the results obtained from the method. The higher variability observed in France is not unexpected because the FR time series covers a larger spectrum of harvest pressure than the GB time series. There is no temporal correlation between the two populations, suggesting that fluctuations of the vital rates are more likely influenced by regional factors than large-scale ones. This is consistent with the literature because both nesting/hatching success and juvenile survival are conditional to the onset of laying, which is tightly related to local weather parameters, e.g. spring temperature, cumulative rainfall, and water levels [@Folliot2017; @Blums2004; @Dzus1998].

The estimates of adult survival rates correspond to values observed in the literature on species of similar weight [@Nichols1997; @Krementz1997; @Buxton2004]. This reflects that there is no apparent scaling issue of this method based on count data. The assumption that male-like individuals correspond only to adult males is then acceptable. This result supports that the count-based method is poorly biased if it is. However, some adult survival rate values are outside the range of expected values even if they are never significantly over 1. A likely reason to these extreme values is a corruption of the closed population assumption on some years. Indeed, if there is an arrival from another Ruddy duck population, the true recruitment cannot explain by itself the population growth and leads to an adult survival combining true adult survival and adult immigration. For the FR population, upper outliers of recruitment rates and survival rates are observed in 2002 and 2012, suggesting immigration events on these years.

The theoretical maximum of the population growth rate is a potential which is determined by intrinsic life history traits of a species [@Dmitriew2010]. This potential is expected to be similar to all Ruddy duck populations anywhere in the world. Realised maximum growth rates that are observed in this study are very close for the two populations. It suggests that the Ruddy duck species reached its intrinsic biological reproduction limit, at least given the environmental conditions of the western Palearctic, i.e. about 0.6 recruits per breeder. The consistency of this intermediate result suggests that the population size estimation is quite accurate for both populations and demonstrates that their changes are well tracked even with a few individuals. The average recruitment rates estimated from the count-based method, with and without harvest, are never higher to the proxies of maximum recruitment rate for both populations. The assumption on the accuracy of the count-based method is then again not excluded from the consistency of this result.

>>>>

---
lang: en-EN
bibliography: "`r paste0('C:/Users/', Sys.info()[['user']], '/Dropbox/JabRef/Library_OFB.bib')`"
csl: https://www.zotero.org/styles/apa # https://osf.io/download/6hw3q/ 
link-citations: true
title: "Exploiting delayed dichromatism to disentangle the effects of adult survival and recruitment on the population dynamics of waterfowl"
author: 
- Adrien Tableau^[Office Français de la Biodiversité, France - adrien.tableau@ofb.gouv.fr]
- Iain Henderson^[Animal and Plant Health Agency, United Kingdom - iain.henderson@apha.gov.uk]
- Sébastien Reeber^[Société Nationale pour la Protection de la Nature, France - sebastien.reeber@snpn.fr]
- Matthieu Guillemain^[Office Français de la Biodiversité, France - matthieu.guillemain@ofb.gouv.fr]
- Jean-François Maillard^[Office Français de la Biodiversité, France - jean-francois.maillard@ofb.gouv.fr]
- Alain Caizergues^[Office Français de la Biodiversité, France - alain.caizergues@ofb.gouv.fr]
date: "`r Sys.setlocale('LC_TIME', 'English'); format(Sys.time(), '%d %B, %Y')`"
abstract: |
  Monitoring the number of individuals is by far the most popular strategy for investigating the environmental factors ruling the population dynamics and for measuring the efficacy of management actions targeting population recovery, control or eradication. Unfortunately, it is insufficient for the basic understanding of the demographic mechanisms and more specifically to assess the extend to which population growth rate is affected by changes in adult survival rather than to variations in reproductive parameters. Usual methods to inform survival and recruitment, e.g. capture-mark-recapture and game-hunting bag, rely on catching animals in a way or another, which suffers from three main drawbacks. First, the selectivity of the catching methods can lead to a biased representation of the underlying population structure. In practice, catching and releasing animals is also an invasive approach implying additional stress to the individuals. Furthermore releasing may simply be forbidden for invasive alien species. In many waterfowl species, a sexual dichromatism is observed in adults whereas immatures of both sexes display a cover-up plumage similar to adult females. From populations of Ruddy duck, a species displaying this pattern, this study introduces a non-invasive method based on count data to estimate the respective contribution of recruitment and adult survival on population growth rate. Relying on population models shows that performing count survey during the appropriate time window to differentiate both plumage types is sufficient to provide long-term time series of the two main demographic components of the population growth rate.

  **Keywords**: reproductive success - fecundity - productivity - age ratio - vital rate - duck - Oxyura jamaicensis
output: 
    bookdown::word_document2:
      toc: false
    bookdown::pdf_document2:
      latex_engine: xelatex
      toc: false
      keep_md: false
      keep_tex: false
header-includes:
  - \usepackage{float} # allows to hold figure position
  - \usepackage[left]{lineno} # line numbering
---

@@

The Ruddy duck is a stiff-tailed duck native from the Americas. From seven individuals initially acclimated in the Slimbridge Wetland Centre in Great Britain in the 40’s [@GutierrezExposito2020], a feral population began to establish with the first observed reproductive attempts in the wild in the 60s (Figure \@ref(fig:count)). This feral population rapidly grew and spread to the entire country to reach more than 5,000 individuals in the early 2000s. By the end of the 80s, a feral population began to establish also on the continent in France, supposedly owing to the arrival of individuals born in Great Britain. Contrary to what was observed in Great Britain, the French Ruddy duck population did not spread much, with the vast majority of observations and breeding attempts clustering in Western France. In France, during winter, almost no Ruddy duck is observed outside the lake of Grand Lieu (47.09°N, 1.67°W), which is thus the exclusive wintering site and greatly facilitates the monitoring of this population. The feral European populations of Ruddy duck are considered as a major threat to the endangered native White-headed duck, *Oxyura leucocephala*, population of the South-Western Mediterranean, because of hybridization risks and thus an elevated potential for genetic pollution and eventually genetic assimilation of the latter by the former [@MunozFuentes2007]. In order to mitigate the risks of genetic pollution of the White-headed duck by Ruddy ducks, eradication measures were taken both in Great Britain and in France from the late 90s [@GutierrezExposito2020], (Figure \@ref(fig:count)), and a European Ruddy duck eradication plan has been initiated by the European Commission in 1999 [@Hughes1999]. The release of non-native species in the wild, including Ruddy ducks, is forbidden in countries of the European Union, and the status of White-headed duck populations is highly unfavourable. The use of capture-mark-recapture to monitor populations of these species is thus impossible. As a result, the effectiveness of the eradication of Ruddy ducks and management actions intended to favour the recovery of White-headed ducks were essentially assessed through censuses. Censuses, however, prevented assessment of the relative effects or changes in survival and recruitment rates on population growth rate, which is a key towards a proper understanding of the efficiency of management actions.

@@

Like many other ducks, Ruddy ducks and White-headed ducks both display delayed dichromatism. Thereby, immature males are confounded to immature and adult females until a first moult occurring in mid-winter for the early ones [@Baldassarre2014], but most likely latter for most of them. Immature males acquire then the typical colourful breeding plumage of their kind in spring [@Baldassarre2014]. Delayed dichromatism explains why the apparent proportion of females in winter counts is always greater than the true proportion of females that can be assessed in counts of late winter or in removals (Figures \@ref(fig:flock) & \@ref(fig:pfem)). Thus, it is possible to assess the proportion of immatures within a population by monitoring the seasonal evolution of apparent sex ratio and assuming an even sex ratio at birth [@Bellrose1961; @Blums1996], comparable mortality rates between females and males among juveniles [@Ramula2018; @Wood2021], and unchanging adult sex ratio over the monitoring period [@Devineau2010]. Then, both adult survival rate and recruitment rate can be estimated by combining the age ratio thus obtained with reliable counts over two successive winters.

@@

In waterfowl, survival rate is commonly higher in adult males than in adult females [@Baldassarre2014]. Adult sex ratio is then generally biased towards males (Figure \@ref(fig:pfem), @Szekely2014). As a consequence, deducing the proportion of immatures within a population from the observed proportion of adult males is not straightforward because the proportion of adults is not just twice as much as the observed proportion of adult males. There were no counts differentiating male-like from female-like individuals during the breeding period to properly estimate adult sex ratio. But sex identification in adults from removals was available for both Ruddy duck populations. However, the small population size in France prevented from getting precise adult sex ratios. A comparison over months in Great Britain showed no difference in proportion of males in adults, so data from adults collected over the whole year were used to estimate the proportion of males in adults. A comparison of proportion of males in adults among years with more than 500 samples did not exhibit significant interannual differences. Even if the adult sex ratio may vary over a long time range, it is relatively stable over a few years [@Wood2021]. As a consequence, removal data from all adults were pooled to estimate the proportion of males in adults, see Equation \@ref(eq:pm).

@@

|Name|Class|Description|
|:---|:----|:----------------------------|
|$AM$|Data|Total number of adult males in removals|
|$AF$|Data|Total number of adult females in removals|
|$p(m|a)$|Parameter|Proportion of males in adults (or probability of being a male knowing it is an adult)|
|$CML_{i, t}$|Data|Cumulated number of male-like individuals counted in population $i$ in year $t$, which are assumed to all be adult males|
|$C_{i, t}$|Data|Cumulated number of individuals counted in population $i$ in year $t$|
|$p(a \cap m)_{i, t}$|Parameter|Proportion of adult males in population $i$ in year $t$|
|$p(a)_{i, t}$|Parameter|Proportion of adults in population $i$ in year $t$|
|$p(i)_{i, t}$|Parameter|Proportion of immatures in population $i$ in year $t$|
|$N_{i, t}$|Data|Size of the population $i$ in year $t$ (maximum number of individuals counted in the wintering population)|
|$s_{i, t}$|Parameter|Adult survival rate, i.e. proportion of individuals in year $t-1$ still alive in year $t$|
|$r_{i, t}$|Parameter|Recruitment rate, i.e. number of recruits in population $i$ in year $t$ per individual in year $t - 1$|
|$S_{i, t}$|Parameter|Number of adults in population $i$ in year $t$|
|$R_{i, t}$|Parameter|Number of recruits in population $i$ in year $t$|

@@

The relevance of the method was assessed by testing the likelihood of the estimates of both vital rates. For adult survival rate, we checked if the values belonged to the expected interval $[0;1]$, and we compared the values to literature data on ducks that are accessible because adult survival is commonly assessed by capture-mark-recapture [@Lebreton2001]. Maximum adult survival is defined as a uniform distribution on $[0.7;1]$. The lower limit corresponds to the upper range of survival rates in literature for waterfowl species of similar weight, the upper limit was set to 1 because survival rates of long-life waterfowl species are very high [@Nichols1997; @Krementz1997; @Buxton2004; @Nichols1992]. Recruitment rate is more tricky to validate because it is not upper bounded as it is defined on $[0;\infty[$ and is neither explicitly informed in literature. Only some components of the recruitment rate are generally described, such as nesting rate, nesting success, clutch size, hatching success, early stage survival, but rarely survival between the juvenile stage and the first breeding season, which is necessary to inform recruitment rate [e.g. @Baldassarre2014]. We then developed an approach to estimate the maximum expected recruitment rate without exploitation and without negative density-dependence processes, and we checked if all values were equal or below this maximum recruitment rate. If the recruitment rate outputs of the model are significantly higher than maximum recruitment rate, it means that the method overestimates such rate, which reflects that a part of adult males are not detected during count surveys. A null or negative recruitment rate would mean that immature males moulted before the survey or/and the assumption on the stability of the adult sex ratio is overrated.

@@

The method successfully provided estimates and associated uncertainty of immature proportion in both wintering populations. Adult survival rate and recruitment rate, i.e. the two components of the population growth rate, were also estimated by the method successfully (Figure \@ref(fig:vital)).

@@

The proportion of males in adults, which is a prerequisite to infer the proportion of immatures, was estimated at 0.60 [0.59; 0.61]. The immature proportion ranged from 0.16 [0.07; 0.24] to 0.54 [0.44; 0.62]. The lower values were similar for the two populations, but the range of the proportions of immatures in the GB population was far more limited (upper value : 0.30 [0.14; 0.45]). As opposed to the FR population, there were no significant differences from a year to another because all estimates overlapped with each other. The proportion of immatures was relatively stable over time in the GB populations, while a slight decrease was observed for the FR population.

Adult survival rate ranged from 0.33 [0.29; 0.37] to 1.01 [0.89; 1.14]. No estimates were significantly outside the range of a survival rates defined without immigration [0; 1]. No trend in adult survival rate was recorded in any of the two populations, although two main differences could be observed. The GB population had the lowest values of adult survival rate, and even if there were significant difference among years for both populations, the variability for the GB population was slightly lower.

Recruitment rate ranges from 0.09 [0.04; 0.13] to 0.75 [0.63; 0.87]. All estimates were above 0, meaning there was no senseless estimation. The maximum recruitment rate being 0.68 [0.36; 0.78] (see Section \@ref(MAX)), no estimates were significantly outside the range of recruitment rates defined without immigration [0; 0.78]. Similarly to the immature proportion, no trend was observed for the GB population, but recruitment rate decreased for the FR population even if it was more noisy than for the immature proportion. As for survival rate, the GB population had low values of recruitment rate, and even if significant differences were observed among years for both populations, the variability for the GB population was far lower than for the FR population.

The lower variability of both vital rates for the GB population explained its steady population growth over the considered time series compared to the FR one (Figure \@ref(fig:count)). In the FR population, the range of recruitment rates (min/max difference: 0.66) was higher than the range of survival rates (min/max difference: 0.58). Conversely, the range of survival rates (min/max difference: 0.46) in the GB population was far higher than the range of recruitment rate (min/max difference: 0.08). This reflected that the variability of the population growth rate for the FR population was mostly driven by recruitment rate whereas the main source of variability was the adult survival for the GB population.

@@

From the five years available in the GB population, proportions of immatures that were estimated from the count-based method were positively correlated to estimates from removal data (Figure \@ref(fig:cor)). The hypothesis that the proportion of female-like individuals in the wintering populations reflected the age structure of these populations was then supported by this result. This correlation was even stronger when considering vital rates (Figure \@ref(fig:cor)). This stronger correlation was expected since the two methods estimating these vital rates had a component in common, the growth rate of the population (see Equation \@ref(eq:vr)).

It was noticeable that a 1:1 correlation was never obtained. Proportion of immatures was lower in counts than in removals. This result was not surprising because immature individuals were expected to be more vulnerable to culling than adults and might then be over represented in removal data. As a consequence, the correlation coefficients for both vital rates also differed from 1. This difference between the two approaches did not question the previous validation of the ability of the count method to track the interannual variability of the vital rates.

@@

## Response of vital rates to culling strategies {#MAX}

Under no culling pressure, maximum growth rates were very similar for the two populations (Figure \@ref(fig:mgr)), respectively 1.45 [1.36; 1.55] and 1.52 [1.25; 1.88] for the GB and the FR population. It corresponded to an increase of 45% per year and 52% per year, respectively. Based on literature, we assumed a maximum survival rate similar for both populations 0.85 [0.7; 1]. We deduced a maximum recruitment rate of 0.60 [0.42; 0.78] for the GB population and of 0.68 [0.36; 1.06] for the FR population.

Even if a significant part of culled individuals were not aged for some years, the proxies of the culling pressure exhibit significant shifts over time (Figure \@ref(fig:sup)). This led to consider a high culling pressure period from 2006 onwards with no clear difference between the two populations because the signal was very noisy. Even if the culling pressure seemed comparable, the culling strategies differed significantly between the two populations, see Section \@ref(HARV). Under high culling pressure, the growth rate dropped to 0.56 [0.53; 0.59] and 0.97 [0.94; 1.00] for the GB and the FR population, respectively, which corresponded to a decrease of 44% per year and 3% per year. This result showed that the GB population significantly decreased during the high culling pressure whereas the FR population was stabilized.

Adult survival rate under high culling pressure was 0.44 [0.42; 0.46] and 0.62 [0.59; 0.64] for the GB and FR populations, respectively, when the recruitment rate decreases to 0.13 [0.11; 0.15] and 0.32 [0.29; 0.34], respectively. The drop in growth rate of the GB population corresponded to a similar average drop in adult survival rate (-0.41) and in recruitment rate (-0.47). The stabilisation of the FR population was caused by a decrease of recruitment rate (-0.36) more than a drop in adult survival (-0.23). Overall, vital rates of the GB population were more affected than the FR population.

@@

The delayed sexual dichromatism in a waterfowl species was used here to develop and test the validity of a new non-invasive method aiming to break down the apparent growth of a population into its two structural components: adult survival rate and recruitment rate. Disentangling the contribution of these two vital rates is a key to understand the response of a population to management measures and potentially adapt these accordingly [@Williams2002]. The development of the method held on a unique dataset of two comparable populations of the same species, the Ruddy duck. Both populations were tracked from their introduction in a similar environment to a period of intense culling pressure. This common history trajectory enabled to evaluate the maximum growth rate of this species, but also its response to culling pressure. As the culling strategy differed between the two populations, we detailed the similarities and the discrepancies of the responses of vital rates, and opened the discussion to the generalization of this estimation method and its implications for waterfowl management.

@@

The vital rate estimates of the two populations belonged to the same range, which demonstrates a certain consistency in the results obtained from the method. The greater variability observed in France is not unexpected because the FR time series covered a larger spectrum of culling pressure than the GB time series. There was no temporal correlation between the two populations, suggesting that fluctuations of the vital rates were more likely influenced by regional factors than large-scale ones. This is consistent with literature because both nesting/hatching success and juvenile survival in waterfowl species are conditional to the onset of laying, which is tightly related to local weather parameters, e.g. spring temperature, cumulative rainfall, and water levels [@Folliot2017; @Blums2004; @Dzus1998].

Count surveys used to test the method occurred in mid-winter, when the first immature males were suspected to moult and start to look like adult males. This might partially corrupt the assumption that male-like individuals corresponded only to adult males. Even though no data on adult survival rate were directly available [@Baldassarre2014], the values estimated from the method developed inhere correspond to what is observed in literature for species of similar weight [@Nichols1997; @Krementz1997; @Buxton2004]. Then, the results reflect that there was no major scaling issue using count data from mid-winter for European Ruddy duck populations. In any case, the time window of the count surveys must be chosen with care, i.e. just before the moult of immature males, to limit the bias of the outputs. On a few years, the values of adult survival rate were outside the range of expected values even if they were never significantly over 1. A likely reason to these extreme values is a corruption of the closed population assumption during some years. Indeed, if there is an arrival from another Ruddy duck population, the true recruitment cannot explain by itself the population growth and leads to an adult survival combining true adult survival and adult immigration. For the FR population, upper outliers of recruitment rates and survival rates were observed in 2002 and 2012, suggesting immigration events during these years.

The theoretical maximum of the population growth rate is a potential which is determined by intrinsic life history traits of a species [@Dmitriew2010; @Niel2005]. This potential is expected to be similar to all Ruddy duck populations anywhere in the world. Realised maximum growth rates that were observed in this study were very close for the two populations. This suggests that the Ruddy duck species reached its intrinsic biological reproduction limit, at least given the environmental conditions of the western Palearctic, i.e. about 0.6 recruits per breeder. The consistency of this intermediate result suggests that the population size estimation was quite accurate for both populations and demonstrates that their changes were well tracked even with relatively few individuals. The average recruitment rates estimated from the count-based method, with and without culling pressure, were never higher to the proxies of maximum recruitment rate for both populations. The assumption on the accuracy of the count-based method is then again not excluded considering the consistency of this result.

Assuming that removal data provided a good picture of the year-to-year variability of age structure of a population, that is, any bias of shooting towards a given age or sex category is relatively stable over time, the strong correlation with the outputs of our method demonstrates its ability to track interannual fluctuations. This result is robust because such correlation was obvious despite being based on a short time series. This strong relationship was also satisfying because it demonstrates that the assumption of a constant adult sex ratio is not corrupted. The temporal autocorrelation of the adult sex ratio might be explained by the fact that it integrates many age-cohorts, which makes it structurally strong. Even if significant variations of the adult sex ratio can still occur over long time periods, this result demonstrates that it is not necessary to monitor and update it on a yearly basis. The adult sex ratio found here on a restricted time period is in line with the Ruddy duck in its native area [0.62 in @Bellrose1980] and with other duck species [@Wood2021]. The count-based method always provided recruitment rates significantly lower than the method based on removal data by a factor close to two, a result that is in line with the hypothesis that the shooting is generally selective towards immature individuals in waterfowl because they are more vulnerable to hunters than adults [@Bellrose1980; @Fox2014]. The gap of vulnerability among ages observed here is realistic because the culling-induced mortality on first-year immatures compared to adults is higher by a factor ranging from 1.3 to 2.6 for hunted duck species in North America [@Bellrose1980].

Overall, even if the true values of vital rates are not known to properly test the accuracy of the count-based method, a beam of arguments supports that the method is not heavily biased. This consists in a big difference with methods based on removal/hunting bag data as demonstrated in @Fox2014, even when one considers an age structure picture that is limited to the end of the hunting season to limit the bias [@Fox2016].
